---
title: "Fixed-shape matrix automatic differentiation"
author: "Sven C. van Dijkman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fixed-shape matrix automatic differentiation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Contract

`ad_matrix_ir()` is a compile-time matrix frontend for the same scalar CppAD
tape engine used by `ADModel`. Every input and intermediate has a fixed shape.
The complete operation graph, shapes, output layout, path contracts, and
lowered scalar IR are serializable; no R matrix calculation occurs during a
recorded value or derivative evaluation.

The supported surface is intentionally statistical rather than arbitrary:
matrix multiplication, transpose, elementwise operations, triangular solves,
SPD and fixed-no-pivot solves, Cholesky/log-determinant, matrix exponential,
symmetric rank updates, quadratic forms, and Cholesky-parameterized covariance
construction. `ad_matrix_supported()` reports the current contract.

## Example

```{r matrix-example}
library(LibeRtAD)

ir <- ad_matrix_ir(
  inputs = list(A = c(2, 2), b = c(2, 1)),
  operations = list(
    ad_matrix_op("L", "cholesky", x = "A"),
    ad_matrix_op("solution", "solve", a = "A", b = "b", method = "spd"),
    ad_matrix_op("logdet", "logdet", x = "A"),
    ad_matrix_op("quadratic", "quadratic_form", a = "A", x = "b")
  ),
  outputs = c("L", "solution", "logdet", "quadratic")
)

point <- list(
  A = matrix(c(4, 1, 1, 3), 2),
  b = matrix(c(2, -1), 2, 1)
)
model <- ad_matrix_compile(ir, at = point, wrt = c("A", "b"))
model$value(point)
model$jacobian(point)
```

Individual matrix elements may be selected with public labels such as
`"b[1,1]"`. A gradient or Hessian requires one scalar output, just as for an
ordinary `ADModel`.

## Path validity

The derivative is exact for the recorded algebraic approximation:

- Cholesky, SPD solve, and `logdet` require a symmetric positive-definite path.
- The general solve is deliberately a fixed no-pivot LU path. If the path is
  poorly conditioned, reformulate or use the SPD route; LibeRtAD does not hide
  pivot changes behind an invalid tape.
- `matrix_exp` uses user-visible fixed scaling and Taylor order. Its derivative
  is exact for that approximation. Accuracy should be checked for the expected
  matrix-norm range.
- Covariance construction can exponentiate the diagonal of a lower-triangular
  factor, ensuring a positive-definite covariance by construction.
- Arbitrary SVD/eigendecomposition and dynamic shapes remain unsupported
  because eigenvector degeneracy, pivot changes, and rank changes do not have
  one globally smooth derivative contract.

Optimized graph caches retain the matrix IR as the source contract and run the
same semantic value/Jacobian probes as scalar caches before reconstruction.
