---
title: "Persistent local and remote job queues"
author: "Sven C. van Dijkman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Persistent local and remote job queues}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Job contract

LibeRties executes LibeRation simulations and estimations through a versioned,
typed job contract. A job contains a serializable `nm_model`, event data,
named entry-point arguments, and an optional label. It must never contain a
live `NMEngine` external pointer.

```{r capabilities}
library(LibeRties)
ls_queue_capabilities()
```

Remote transport uses typed JSON. Model expression IR is recompiled from
semantic fields by the receiving host; the server does not deserialize an RDS
payload supplied over HTTP and does not expose an arbitrary-code endpoint.

## Create a local queue

```{r queue}
queue <- ls_local_queue(
  root = tempfile("liberties-queue-"),
  user = "local",
  max_workers = 1L,
  limits = list(
    max_runtime_seconds = 3600,
    max_memory_mb = 4096,
    max_queued_jobs = 100
  )
)
queue
queue$list()
```

The queue root contains durable job metadata, checksums, payloads, results,
and log streams. Reopening the same root recovers terminal history and checks
whether previously running worker processes are still alive.

## Submit and monitor work

```{r submit, eval=FALSE}
job <- ls_job(
  "estimate",
  model = model,
  data = data,
  arguments = list(
    method = "FOCEI",
    covariance = TRUE,
    print_every = 10,
    n_cores = 2
  ),
  label = "FOCEI base model"
)

id <- queue$submit(job)
queue$status(id)
queue$logs(id, "stdout")
final <- queue$wait(id, timeout = 3600)

if (identical(final$status, "completed")) {
  fit <- queue$result(id)
}
```

States are `queued`, `running`, `completed`, `failed`, and `cancelled`.
`$poll()` reaps workers, enforces limits, recovers untracked jobs, and starts
queued work when capacity is available. `$cancel()` terminates tracked work
and durably records cancellation.

## Integrity and provenance

Payload and result SHA-256 digests are recorded in metadata and verified on
read. Each worker runs in a fresh R process with a scrubbed environment and a
per-job working directory. Runtime, CPU, memory, result-size, queue, and
storage limits are applied at the queue boundary.

The default subprocess queue is application-level isolation for trusted local
use. A production server instead passes `ls_systemd_executor()` to the queue:
each job becomes a transient systemd user service with namespace isolation and
whole-cgroup CPU, task, memory, and wall-time limits. A request for four cores
receives a 400% CPU quota, so PSOCK/FORK parallelism is retained inside the
single job cgroup.

On an HPC submission host, pass `ls_slurm_executor()` or
`ls_grid_engine_executor()` instead. The scheduler receives native core,
whole-job memory, and wall-time requests, while the existing LibeRties job
directory remains the durable exchange boundary for payloads, logs, results,
and metadata. Scheduler IDs are persisted and reattached after an API restart.

```{r scheduler, eval=FALSE}
executor <- ls_grid_engine_executor(
  parallel_environment = "smp",
  memory_resource = "mem",
  runtime_resource = "h_rt",
  max_cores_per_job = 36L
)
queue <- ls_local_queue(
  "~/Scratch/LibeRties", max_workers = 20L, executor = executor
)
```

The root and installed R libraries must be visible from compute nodes. A
scheduler enforces resources but does not automatically prove tenant
filesystem or network isolation; production use needs an independently
attested sandbox.

For a personal scheduler account, LibeRation can use the same executors without
an HTTP service. The **Direct SSH scheduler** route calls the fixed
`ls_direct_scheduler_cli()` entry point over OpenSSH, optionally via a jump
host. It transmits only the versioned typed JSON job contract and stores the
queue and its encryption credential under the remote account. Scheduler IDs,
logs and results therefore remain recoverable after a client disconnect. This
route inherits the SSH account's permissions and is distinct from the
centrally administered multi-user API deployment.

## NONMEM and nlmixr2 workers

Simulation and estimation jobs can use an allow-listed external engine while
retaining the same durable queue and result-retrieval path.

```{r external-engine-job, eval=FALSE}
job <- ls_job(
  "estimate", model, data, engine = "nonmem",
  arguments = list(
    method = "FOCEI", maxit = 500L, covariance = TRUE,
    audit_artifacts = TRUE
  )
)
id <- queue$submit(job)
```

NONMEM/PsN and nlmixr2/rxode2 are installed and configured by the worker
operator. No executable path, raw R function, shell fragment, module command,
or scheduler option is accepted in the submitted job. `ls_queue_capabilities()`
reports local engine discovery; `remote$capabilities()` reports the API host.
When a systemd sandbox hides a licensed installation below a home directory,
the operator may expose that path read-only through
`ls_systemd_executor(external_read_paths = ...)`. Scheduler deployments load
licensed software in their trusted `prologue`.

## Connect to a remote service

```{r remote, eval=FALSE}
remote <- ls_remote(
  "https://liberties.example.org",
  token = Sys.getenv("LIBERTIES_TOKEN")
)

remote$authenticate()
id <- remote$submit(job)
remote$status(id)
remote$logs(id, "stdout")
fit <- remote$result(id)
```

The remote client mirrors the local operations closely. HTTPS is strongly
recommended outside a trusted private network because bearer tokens authorize
all user operations.
