---
title: "Optimal trial design with LibeRality"
author: "Sven C. van Dijkman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Optimal trial design with LibeRality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(LibeRality)
```

LibeRality connects a clinically recognisable trial description to the same
NONMEM-compatible model and C++ numerical engine used by LibeRation. A design
contains arm schedules, endpoint definitions, population strata, uncertainty
scenarios, design variables, constraints, and optional prior information.

## Build or open a design

The included example is an oral population-PK study with rich and sparse arms.

```{r example}
example <- lity_example()
design <- example$design
design
lity_validate(design)
```

`lity_schedule()` uses familiar event columns (`TIME`, `EVID`, `AMT`, `RATE`,
`II`, `ADDL`, `SS`, `CMT`, and `DVID`). This allows doses, infusions, repeated
doses, steady state, covariates, and multiple outcomes to retain their
LibeRation/NONMEM meaning.

## Information and precision

```{r information}
information <- lity_information(design)
information
head(data.frame(
  parameter = names(information$se),
  se = information$se,
  rse = information$rse
))
```

For continuous observations, LibeRality uses the multivariate-normal
population expected information, including parameter effects on the mean and
the observation covariance. Prediction derivatives come from LibeRation's
CppAD-backed engine. OMEGA derivatives are analytic; residual covariance
derivatives retain the declared error model. Binary, ordinal, count, recurrent,
and event-time designs use distribution-aware counting-process or working-
moment information.

## Evaluate competing objectives

```{r criteria}
evaluation <- lity_evaluate(design, list(
  local_D = lity_criterion_D(),
  robust_D = lity_criterion_robust(lity_criterion_D()),
  max_rse = lity_criterion_rse(summary = "max"),
  cost = lity_criterion_cost(),
  burden = lity_criterion_burden()
))
evaluation
```

Every criterion declares whether it is maximised or minimised. Compound
criteria require explicit weights and reference scales. `lity_criterion_pareto()`
keeps objectives separate and `lity_pareto()` returns the non-dominated set.

## Optimise and verify

The example mixes sampling-time and integer arm-size variables, so the hybrid
algorithm is appropriate. The shortened run below is illustrative; real work
should use convergence checks, multiple seeds, and adequate iteration limits.

```{r optimise, eval=FALSE}
criterion <- lity_criterion_compound(
  list(information = lity_criterion_D(), cost = lity_criterion_cost()),
  weights = c(0.8, 0.2),
  reference = c(1, 100000)
)
result <- lity_optimise(
  design, criterion, method = "hybrid",
  control = list(maxit = 60, particles = 30)
)

verification <- lity_simulate_trials(result$design, n = 250)
lity_report(result, "optimal-design-report.html")
```

The empirical simulation can include dropout, non-adherence, missed samples,
random effects, residual error, and scenario sampling. Set `fit = TRUE` to
re-estimate each trial with a chosen LibeRation algorithm; this is substantially
more expensive but provides empirical bias, RMSE, convergence, and coverage
checks against theoretical precision.

## Ecosystem hand-off

- `lity_model_from_liberary()` imports a reviewed catalogue model.
- `lity_model_from_liberation()` imports either the initial values from a
  model version or the final values from a completed estimation run.
- `liberality_gui()` presents both sources, built-in structural/likelihood
  templates, and local NONMEM control streams through a common preview and
  compatibility workflow. The selected model is copied into the design with a
  hash and provenance rather than linked to mutable external state.
- `lity_endpoint_from_liberator()` preserves a therapeutic endpoint definition.
- `lity_to_liberation()` produces an exact event template or simulated dataset.
- `lity_job()` sends evaluation, optimisation, simulation, or Pareto work to a
  LibeRties queue.

Use `liberality_gui()` for the React workbench. Long optimisations should be
queued, checkpointed, and accompanied by scenario and simulation sensitivity
analyses.
